\documentclass[11  pt]{article} 
\usepackage[lmargin=1in,rmargin=1.75in,bmargin=1in,tmargin=1in]{geometry}  


\input{preamble}

\begin{document}
	
	\lecture{: Approximation Algorithms and Optimization}{Week 14}
	\textbf{Course Logistics}
	\begin{itemize}
		\item CLRS Chapter 29 and 35
		\item Last homework due Friday
	\end{itemize}


	
	\section{Approximation Algorithms for Optimization Problems}
	There are so many hard problems out there without known polynomial time solutions! Is all hope lost? What do we do?\\
	
	Approximation algorithms ``quickly'' find an answer that is ``close'' to the solution. \\ \\
	
	\vs{3cm} 
	
	We are now focused on optimization problems, rather than just their \emph{decision} versions.
	\paragraph{Definition}
	Let $Q$ be a computational minimization problem (e.g., find a minimum vertex cover, find a minimum $s$-$t$ cut) and assume that $C^*$ is the optimal (minimum) solution to the problem. An \hide{approx algorithm} for $Q$ with approximation factor $p$ is an algorithm that 
	\begin{itemize}
		\item Runs in %polynomial time
		\item Outputs a solution value $C$ that is guaranteed to satisfy % $\frac{C}{C^*} \leq p$.\\
	\end{itemize}
	
	\newpage
	
	\begin{Qu}
		What is a lower bound we must assume holds for the value of $p$ when defining an approximation algorithm?
		\begin{itemize}
			\aitem No lower bound is needed, $p$ can be any real number
			\bitem $p \geq -1$
			\citem $p \geq 0$
			\ditem $p \geq 1$
			\eitem $p \geq C^*$
		\end{itemize}
	\end{Qu}
	\vs{1cm}
	
	We can also have approximation algorithms for maximization problems, but in this case $C^*$ represents the \emph{maximum} value (i.e., the solution) for the problem and \hide{$p \leq 1$.} %We will focus on the minimization case here.\\ \\
	
	
	
	\newpage
	
	\paragraph{Wait, is this even possible?}
	We want to guarantee that $\frac{C}{C^*}$ is small in polynomial time, but finding $C^*$ is NP-hard. How can we do that? \\
	
	To design an approximation algorithm, we need two pieces:
	\begin{enumerate}
		\item \hide{Lower bound.} A procedure for finding a value $\hat{C}$ that satisfies \hide{$\hat{C} \leq C^*$}
		\begin{itemize}
			\item It should take polynomial time \\
			\item This method \hide{does not} solve the original problem, but is often related
		\end{itemize}
		\item \hide{Upper bound.} An algorithm for the original problem that returns \\
		
		a suboptimal solution \hide{$C \geq C^*$} \\ \\
		\begin{itemize}
			\item It should take polynomial time \\
			\item This method often: \\
		\end{itemize}
	\end{enumerate}
	
	We then must prove that \hide{$\frac{C}{\hat{C}} \leq p$} and this provides a $p$-approximation! 
	
	\vfill
	
	
	\textbf{Caveat.} Often, the lower bounding procedure is \emph{explicit}: there is an actual algorithm that computes the lower bound, and then the upper bounding algorithm explicitly uses the lower bound. However, in some cases no explicit lower bound is computed, and instead one shows implicitly that the upper bounding algorithm has to provide a solution that is better than some lower bound, even though the lower bound isn't computed explicitly. This can be tricky, but it is often possible!
	
	\newpage
	\section{Matchings and Vertex Covers}
	Let $G = (V,E)$ be an undirected and unweighted graph. \\
	
	A matching $\mathcal{M} \subseteq E$ is a set of edges such that no two edges share the same node. \\ 
	
	A vertex cover $\mathcal{C} \subseteq V$ is a set of nodes such that each edge $(u,v) \in E$ has at least one node in $\mathcal{C} \subseteq V$.\\
	\vspace{5cm} 
	
	
	\begin{lemma}
		Let $\mathcal{M}$ be a matching of $G$ and $\mathcal{C}$ be a vertex cover. Then $|\mathcal{M}| \leq |\mathcal{C}|$.
	\end{lemma}

	
	%Assume that is not the case, and that $\mathcal{M}$ has more edges than $\mathcal{C}$ has nodes.  \\
	%	
	%	Let $S$ be the set of nodes adjacent to an edge in $\mathcal{M}$. Since all of these nodes touch exactly one edge in $\mathcal{M}$, we know that $|\mathcal{M}| = \frac{|S|}{2}$.
	

	
	\newpage
	\section{The approximation algorithm for vertex cover}
	A matching $\mathcal{M} \subseteq E$ is a \emph{maximal} matching if for every edge $e \in E - \mathcal{M}$, $\mathcal{M} \cup \{e\}$ is no longer a matching. \\
	
	
	\vspace{8cm}
	
	
	\textbf{The algorithm}\\
	
	$\textsc{VertexCoverApprox}(G = (V,E))$
	\begin{enumerate}
		\item Compute a maximal matching $\mathcal{M}$ of $G$:
		\begin{itemize}
			\item Set $F = E$, $\mathcal{M} = \emptyset$
			\item While $|F| > 0$
			\begin{itemize}
				\item Add any edge $e \in F$ to $\mathcal{M}$
				\item For each remaining $f \in F$, if $|e \cap f| > 0$, remove $f$ from $F$
			\end{itemize}
		\end{itemize}
		\item Let $S$ be the set of nodes adjacent to an edge in $\mathcal{M}$
		\item Return $S$\\
	\end{enumerate}
	
	\newpage
	\begin{theorem}
		Let $C^*$ be the minimum sized vertex cover of $G$. The algorithm \textsc{VertexCoverApprox} runs in polynomial time in terms of the size of $G$ and outputs a vertex cover $S \subseteq V$ satisfying $|S| \leq 2C^*$. Thus, this is a 2-approximation algorithm for vertex cover.
	\end{theorem}


	




\section{Graph problems as mathematical optimization problem}
We can write the maximum $s$-$t$ flow problem as the following optimization problem:


\vfill

The weighted vertex cover problem can be written as the following optimization problem:
\vfill

\begin{Qu}
	Which of the above two problems is a linear program?
	\begin{itemize}
		\aitem The first
		\bitem The second
		\citem Both
		\ditem Neither
	\end{itemize}
\end{Qu}


\section{Linear Programming Relaxation for Weighted Vertex Cover}
This is the integer program for the weighted vertex cover problem:



\vfill

\begin{Qu}
	Let $\hat{C}$ be the optimal solution value to the linear programming relaxation of the weighted vertex cover integer program, and let $\mathcal{C}^*$ be the optimal solution to the weighted vertex cover problem. Which of the following is always true?
	\begin{itemize}
		\aitem $\hat{C} \leq C^*$
		\bitem $\hat{C} \geq C^*$
		\citem $\hat{C} < C^*$
		\ditem $\hat{C} > C^*$
		\eitem $\hat{C} = C^*$
	\end{itemize}
\end{Qu}

\newpage
\section{The Algorithm}
	$\textsc{WeightedVertexCoverApprox}(G = (V,E))$
\begin{enumerate}
	\item Solve the linear programming relaxation of the weighted vertex cover problem
	\item For each node $v \in V$, if $x_v \geq 1/2$, add $v$ to a node set $S$. In other words, define
	\begin{equation*}
		S = \{v \in V \colon x_v \geq 1/2\}
	\end{equation*}
	\item Return $S$ as a vertex cover
\end{enumerate}
\begin{theorem}
	Let $C^*$ be the weight of the minimum weighted vertex cover of $G$. The algorithm \textsc{WeightedVertexCoverApprox} runs in polynomial time in terms of the size of $G$ and outputs a vertex cover $S \subseteq V$ satisfying $\sum_{v \in S} w_v\leq 2C^*$. 
\end{theorem}
	
	
	
	
	
	
	
\end{document}